Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2020-04-02 treestats_paper.tex
edits for JK comments

Population genetic data is most directly represented as a matrix of genotypes,
and so summaries of popluation genetic data usually begin with this data structure.
However, since the genetic variation represented in a genotype matrix
was generated by mutations inherited down genealogical trees,
it is often informative to think about these statistics in terms of these trees.
Most population genetic summary statistics can be thought of in terms of the \emph{allele frequencies},
i.e., what proportion of the sampled genotypes carry each allele.
The frequency of any particular allele
can easily be found from the location of the mutation(s) that produced it
on the marginal genealogical tree,
by counting the number of sampled genotypes below it in the tree.
If many mutations have occurred on a single genealogical tree,
then these mutations start to give us an idea of the shape of that tree,
and the frequencies of all of them can be easily found by propagating sample counts up the tree.
This relationship is depicted in \autoref{fig:allele_freqs},
which shows the relationship between the genealogical tree
and the frequencies of alleles at five different sites across a region of the genome
described by that single tree.
Our goal will be to generalize this sort of summary of the genotype matrix,
using the underlying genealogical trees.
Working with the trees serves two purposes.
First, the trees are closer to the demographic processes that we are really interested in,
and so allow us to reason more directly about population genetic signal and noise.
Second, trees are highly efficient data structures,
and allow us to compute with genetic data far more efficiently than with the genotype data itself.

\begin{figure}
    \begin{center}
        \includegraphics{figures/allele_frequency_diagram}
    \end{center}
    \caption{
        The relationship between allele frequencies and the underlying genealogical trees.
        \textbf{(a)} The genotype matrix, showing genotypes of five sequences at five separate sites,
        as well as the derived allele frequencies (derived alleles are shown in color).
        \textbf{(b)} The genealogical tree describing how those five samples are related to each other
        over this segment of genome. The samples are leaves,
        and each mutation is represented as a star on the genealogy,
        with color matching the genotype matrix, and
        labeled with its index in the genotype matrix
        and the ancestral and derived states at that site.
        \textbf{(c)} Each node in the tree is labeled with the number of samples
        at or below that node, which we think of as a \emph{weight}.
        Note that this is \emph{additive}:
        the weight of each node is the sum of the weights of its child nodes.
        Since the frequency of a derived allele is the number of samples
        that have inherited that mutation, the allele frequency of each mutation
        is equal to the weight of the node directly below the mutation,
        divided by the total sample size.
        \label{fig:allele_freqs}
    }
\end{figure}

The operation of finding allele frequencies on a tree,
depicted in \autoref{fig:allele_freqs}(c)
is the basis of a familiar operation in population genetics:
summarizing the allele frequency spectrum.
Somewhat more generally, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other samples weight zero), then
(b) finding the total weight of all samples in the subtree inheriting from that branch.
This gives the number of samples that would inherit any mutation falling on this branch.
Suppose we want to calculate one of the linear functions of the frequency spectrum
discussed by \citet{fu1995statistical} and \citet{achaz2009frequency},
which can be described as follows.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci and all alleles
(including the ancestral allele),
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
For instance, mean genetic diversity across biallelic loci is computed using this formula
with $f(p) = p (1-p)$,
omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
This is because genetic diversity is defined to be the probability that two randomly chosen sequences
differ at a randomly chosen site,
and the probability that two randomly chosen genomes differ at a site with allele frequency $p$
is $p (1-p) + (1-p) p$.

>>>>

Population genetic data is most directly represented as a matrix of genotypes,
and so summaries of population genetic data usually begin with this data structure.
However, since the genetic variation represented in a genotype matrix
was generated by mutations inherited through genealogical trees,
it is often informative to think about these statistics in terms of these trees.
Most population genetic summary statistics can be thought of in terms of the \emph{allele frequencies},
i.e., what proportion of the sampled genotypes carry each allele.
The frequency of any particular allele
can easily be found from the location of the mutation(s) that produced it
on the marginal genealogical tree,
by counting the number of sampled genotypes below it in the tree.
If many mutations have occurred on a single genealogical tree,
then these mutations start to give us an idea of the shape of that tree.
This relationship is depicted in \autoref{fig:allele_freqs},
which shows the relationship between the genealogical tree
and the frequencies of alleles at five different sites across a region of the genome
described by that single tree.
Our goal will be to generalize this sort of summary of the genotype matrix,
using the underlying genealogical trees.
Working with the trees serves two purposes.
First, the trees are closer to the demographic processes that we are really interested in,
and so allow us to reason more directly about population genetic signal and noise.
Second, trees are highly efficient data structures,
and allow us to compute with genetic data far more efficiently than with the genotype matrix itself.

\begin{figure}
    \begin{center}
        \includegraphics{figures/allele_frequency_diagram}
    \end{center}
    \caption{
        The relationship between allele frequencies and the underlying genealogical trees.
        \textbf{(a)} The genotype matrix, showing genotypes of five sequences at five separate sites,
        as well as the derived allele frequencies (derived alleles are shown in color,
        with separate colors for each genotype pattern).
        \textbf{(b)} The genealogical tree describing how those five samples are related to each other
        over this segment of genome. The samples are leaves,
        and each mutation is represented as a star on the genealogy,
        with color matching the genotype matrix, and
        labeled with its index in the genotype matrix
        and the ancestral and derived states at that site.
        \textbf{(c)} Each node in the tree is labeled with the number of samples
        at or below that node, which we think of as a \emph{weight}.
        Note that this is \emph{additive}:
        the weight of each node is the sum of the weights of its child nodes.
        Since the frequency of a derived allele is the number of samples
        that have inherited that mutation, the allele frequency of each mutation
        is equal to the weight of the node directly below the mutation,
        divided by the total sample size.
        \label{fig:allele_freqs}
    }
\end{figure}

The operation of finding allele frequencies on a tree,
depicted in \autoref{fig:allele_freqs}(c)
is the basis of a familiar operation in population genetics:
summarizing the allele frequency spectrum.
Somewhat more generally, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other samples weight zero), then
(b) finding the total weight of all samples in the subtree inheriting from that branch.
This gives the number of samples that would inherit any mutation falling on this branch.
Suppose we want to calculate one of the linear functions of the frequency spectrum
discussed by \citet{fu1995statistical} and \citet{achaz2009frequency},
which can be described as follows.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple sum over sites,
we can compute this by summing over the $L$ sites in the genome and all alleles
(including the ancestral allele),
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
For instance, mean genetic diversity across biallelic loci is computed using this formula
with $f(p) = p (1-p)$,
omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
This is because genetic diversity is defined to be the probability that two randomly chosen sequences
differ at a randomly chosen site,
and the probability that two randomly chosen genomes differ at a site with allele frequency $p$
is $p (1-p) + (1-p) p$.
